---
title: "Hilton International Hotels Case Study"
author: "Teresiah Karumba"
date: "2023-07-07"
output:
  pdf_document: default
  html_document: default
---

### Loading required packages 
```{r}

pacman::p_load(tidyverse, dplyr, openxlsx, janitor, ggplot2, lubridate,latexpdf)

```


### Working Directory 
```{r}

setwd("C:/Users/teresiah.karumba/Burn Manufacturing/Marketing - Business Intelligence/BI Automations 2020/Teresiah/Coding Challenges/")

getwd()

```

### Loading the Dataset 
```{r}

hotel_df <- read.csv("hotel_bookings.csv (1)/hotel_bookings.csv") %>% clean_names()
names(hotel_df)

```

### Data Structure 
```{r}
# Data structure

class(hotel_df) #returns the class attribute of our data

glimpse(hotel_df) #checking the data types of our columns

head(hotel_df) #Having a view of our data set.

dim(hotel_df) #shape of the data #119390 rows of data #32 rows of data

summary(hotel_df) #Summary of our data set


#character columns to factor

hotel_df <- hotel_df %>% mutate(hotel = as.factor(hotel),
                     meal = as.factor(meal),
                     country = as.factor(country),
                     market_segment = as.factor(market_segment),
                     distribution_channel = as.factor(distribution_channel),
                     reserved_room_type = as.factor(reserved_room_type),
                     assigned_room_type = as.factor(assigned_room_type),
                     deposit_type = as.factor(deposit_type),
                     customer_type = as.factor(customer_type),
                     reservation_status = as.factor(reservation_status),
                     is_canceled = as.factor(is_canceled),
                     is_repeated_guest = as.factor(is_repeated_guest),
                     arrival_date_month = as.factor(arrival_date_month)
                    )

#Attaching levels to the factor variables
hotel_df$is_canceled <- factor(hotel_df$is_canceled, labels = c("No","Yes"))

hotel_df$is_repeated_guest <- factor(hotel_df$is_repeated_guest, labels = c("No","Yes"))


#Merging Columns

# hotel_df <- hotel_df %>% unite(arrival_date, arrival_date_year,arrival_date_month,arrival_date_day_of_month,sep = "-")


```

### Data Cleaning 
##### Data Completeness
```{r}
listMissingColumns <- colnames(hotel_df)[ apply(hotel_df, 2, anyNA)]
print(listMissingColumns) #Children column has missing data

hotel_df %>% filter(is.na(children)) -> miss_children

#omit any data with NA

na.omit(hotel_df)-> hotel_df

dim(hotel_df) #shape of the data #119386 rows of data #32 rows of data

```

### Data Analysis 
##### How many bookings were cancelled? 
```{r}

#Creating the proportions

cancelled_bookings <- hotel_df %>% 
  group_by(is_canceled) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n`/sum(`n`)) %>% 
  mutate(labels = scales::percent(perc))


print(cancelled_bookings)


plot1 <- ggplot(cancelled_bookings, aes(x = "", y = perc, fill = is_canceled)) +
  geom_col(color = "black") +
  geom_label(aes(label = labels), colour = c("white", 1),
            position = position_stack(vjust = 0.5),
            show.legend = FALSE) +
  guides(fill = guide_legend(title = "Number of Bookings Cancelled")) +
  scale_fill_viridis_d() +
  coord_polar(theta = "y") + 
  theme_void()


print(plot1)


```

### What was the booking ratio between resort hotels and city hotels? 
```{r}
#Creating the proportions

resort <- hotel_df %>% filter(hotel == "Resort Hotel")
city <- hotel_df %>% filter(hotel == "City Hotel")

resort_count <- nrow(resort)
city_count <- nrow(city)

ratio <- resort_count/city_count

# Print the ratio in ratio format
ratio_string <- sprintf("%d:%d", resort_count, city_count)

print(ratio_string)

print(ratio)



hotel_ratio <- hotel_df %>% 
  group_by(hotel) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n`/sum(`n`)) %>% 
  mutate(labels = scales::percent(perc))



plot2 <- ggplot(hotel_ratio, aes(x = "", y = perc, fill = hotel)) +
  geom_col(color = "black") +
  geom_label(aes(label = labels), colour = c("white", 1),
            position = position_stack(vjust = 0.5),
            show.legend = FALSE) +
  guides(fill = guide_legend(title = "Number of Bookings in per Hotel")) +
  scale_fill_viridis_d() +
  coord_polar(theta = "y") + 
  theme_void()


print(plot2)

```

### What was the percentage booking for each year? 
```{r}
#Creating the proportions

t1 <- table(hotel_df$arrival_date_year)

t2 <- round(prop.table(t1)*100, digits = 2)

print(t2)

# Data frame with the percentages

booking_yrdf <- as.data.frame(t2)


names(booking_yrdf) <- c("Year", "Percentage")


booking_yrdf <-  booking_yrdf[order(booking_yrdf$Percentage, decreasing = TRUE),]

plot3 <- ggplot(booking_yrdf, aes(x = Year, y = Percentage)) +
  geom_bar(stat = "identity", mapping = aes(x = Year, fill = Year)) +
  geom_text(aes(label = Percentage), vjust = 0, colour = "black") +
  scale_fill_viridis_d()


#adding Chart Title
plot3 <- plot3 + ggtitle("Percentage Bookings per Year") + theme(plot.title = element_text(hjust = 0.5))

plot3 <- plot3 + labs(y="No of Bookings") 

plot3 <- plot3 + labs(x="Year") 


print(plot3)

```


### Which was the busiest month for hotels?
```{r}
#August was the busiest month for the hotels

t3 <- table(hotel_df$arrival_date_month)

# Data frame with the counts

busy_month <- as.data.frame(t3)


names(busy_month) <- c("Month", "Tally")


busy_month <-  busy_month[order(busy_month$Tally, decreasing = TRUE),]


plot4 <- ggplot(busy_month, aes(x = Month, y = Tally)) +
  geom_bar(stat = "identity", mapping = aes(x = Month, fill = Month)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()


#adding Chart Title
plot4 <- plot4 + ggtitle("Bookings per Month") + theme(plot.title = element_text(hjust = 0.5))

plot4 <- plot4 + labs(y="No of Bookings") 

plot4 <- plot4 + labs(x="Month") 



print(plot4)

```


### Most Guest come from which Country? 
```{r}
#Most guests came from Portugal (PRT)

t4 <- table(hotel_df$country)

guest_df <- as.data.frame(t4)

names(guest_df) <- c("Country", "Count")

guest_df <-  guest_df[order(guest_df$Count, decreasing = TRUE),]

top_10_countries <- head(guest_df, n=10)


plot5 <- ggplot(top_10_countries, aes(x = reorder(Country, - Count) , y = Count)) +
  geom_bar(stat = "identity", mapping = aes(x = Country, fill = Country)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_fill_viridis_d()

#adding Chart Title
plot5 <- plot5 + ggtitle("Top 10 Countries with most Guests") + theme(plot.title = element_text(hjust = 0.5))

plot5 <- plot5 + labs(y="No of Guests") 

plot5 <- plot5 + labs(x="Country") 


print(plot5)



```

### How long do most people stay in hotels?
```{r}

#Creating a column to calculate total number of days stayed

hotel_df <- hotel_df %>% 
  mutate(total_days = stays_in_weekend_nights + stays_in_week_nights)


avg_stay <- mean(hotel_df$total_days) %>% round(0) %>% as.character()

print(paste("Average stay at the hotel is", avg_stay, "days"))


med_stay <- median(hotel_df$total_days) %>% as.character()

print(paste("Most people stay at the hotel for", med_stay, "days"))



plot6 <- ggplot(hotel_df, aes(x=total_days,fill=hotel)) + 
  geom_histogram(binwidth=1, alpha=0.5, position = 'identity') +
  labs(title  = "Distribution of No. of stays",
       fill = "hotel") 
  scale_fill_viridis_d()

print(plot6)

```

### The most booked accomodation type
```{r}
t5 <- table(hotel_df$customer_type)

accomodation_df <- as.data.frame(t5)

names(accomodation_df) <- c("Customer_Type", "Count")

accomodation_df[order(accomodation_df$Count, decreasing = TRUE),]


plot7 <- ggplot(accomodation_df, aes(x = Customer_Type , y = Count)) +
  geom_bar(stat = "identity", mapping = aes(x = Customer_Type, fill = Customer_Type)) +
   scale_fill_viridis_d()

#adding Chart Title
plot7 <- plot7 + ggtitle("Customer Type Distribution") + theme(plot.title = element_text(hjust = 0.5))

plot7 <- plot7 + labs(y="No of Guests") 

plot7 <- plot7 + labs(x="Customer Type") 


print(plot7)


```
